---
title: "Out-of-Sample Cross-Validation of the RCV Support Models"
author: "Jesse Brandt"
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(tidyr)
library(RColorBrewer)

load("../data/cv_results.RData")   # cv_results (from R/05_cv.R)

metrics      <- cv_results$metrics
fold_metrics <- cv_results$fold_metrics
fold_mean    <- cv_results$fold_mean
imp_spread   <- cv_results$imp_spread
prov         <- cv_results$provenance

# convenience: a named getter for headline numbers used in prose
m1 <- metrics[1, ]; m2 <- metrics[2, ]; m3 <- metrics[3, ]
pp <- function(x) sprintf("%.1f", 100 * x)   # proportion -> percentage-point string
```

## Why this report exists

The models in `04_models.R` report an in-sample adjusted pseudo-R² of **0.86**
(baseline) and **0.83** (+ demographics). Those numbers measure fit to the *same*
precincts the models were trained on. They cannot tell a reader how well the model
would predict RCV support in a **jurisdiction it has never seen** — which is the
only claim worth putting on a résumé. This report replaces the in-sample number
with a genuinely out-of-sample one.

> **Headline:** under leave-one-locale-out cross-validation, the baseline model
> `dem_share * recent_lpw` predicts a held-out jurisdiction's precinct-level RCV
> support with a mean absolute error of **`r pp(m1$mae)` percentage points**
> (out-of-sample deviance-R² = **`r sprintf("%+.2f", m1$oos_dev_r2)`**), pooling
> across `r prov$m_imputations` imputations. Every number below is out-of-sample.

## Design decisions

**Grouping: leave-one-locale-out (LOLO), `r prov$n_folds` folds.**
Precincts are nested within `r prov$n_folds` locales, and locale drives almost
everything — the outcome, the key regressor `recent_lpw` (a locale-level
constant), and M3's random intercept. A naive random precinct split would let the
model learn each locale's mean from its own training precincts and then "predict"
its held-out precincts, leaking exactly the cluster structure we want to test
across. Each fold therefore holds out one **entire** locale and predicts it from
the other seven. This is the correct — and deliberately harsh — test of
generalisation to a new jurisdiction. It can honestly produce a **negative**
deviance-R² when a model does worse than a training-mean intercept.

Only Alaska and Maine have `recent_lpw = TRUE`, so LOLO keeps `recent_lpw`
identified in every fold (holding out one leaves the other). Holding out Alaska or
Maine is thus a direct out-of-sample test of the paper's central LPW hypothesis:
*does the LPW shift estimated from Maine alone actually predict Alaska, and vice
versa?*

**Imputation under CV (leakage-aware).**

- **M1** uses only `dem_share * recent_lpw` — no demographics, so **no imputation**.
  Its CV is fully leakage-free. This is the cleanest headline number.
- **M2 / M3** use six Census demographics with missing values. The `mice`
  imputation model is **refit inside each training fold** via mice's `ignore=`
  argument: the held-out locale's rows are imputed but never contribute to
  estimating the imputation model, so no held-out information leaks into the
  imputed predictors. *Documented compromise:* under LOLO the held-out locale is
  an unseen factor level, so we drop `locale` from the imputation predictors
  (03_imputation.R used it) and impute demographics from `dem_share`,
  `recent_lpw`, and the other demographics. Held-out imputation therefore regresses
  toward training patterns — the honest reality of predicting a new jurisdiction's
  missing demographics.

**Metrics** (appropriate to a quasibinomial / proportion outcome):

- **OOS deviance-explained** — `1 − Σ deviance(model) / Σ deviance(null)`, null =
  each fold's training-set overall support rate, pooled over all held-out
  precincts. The out-of-sample analog of the in-sample pseudo-R².
- **MAE / RMSE** of the predicted support share (percentage points), precinct-level
  and vote-weighted.

All metrics pool across the `r prov$m_imputations` imputations by averaging the
predicted probabilities per precinct, then scoring once.

## Pooled out-of-sample results

```{r pooled-table}
metrics %>%
  transmute(
    Model = model,
    `OOS dev-R²` = round(oos_dev_r2, 3),
    `MAE (pp)`   = round(100 * mae, 2),
    `RMSE (pp)`  = round(100 * rmse, 2),
    `Vote-wt MAE (pp)` = round(100 * mae_vote_wt, 2),
    `N held-out precincts` = n_precincts
  ) %>%
  kable(caption = paste0("Leave-one-locale-out metrics, pooled across ",
                         prov$m_imputations, " imputations. Every precinct scored ",
                         "only when its locale was held out.")) %>%
  kable_styling(full_width = FALSE)
```

Two ways of averaging tell slightly different stories: **precinct-pooled** (above,
dominated by the large Massachusetts and Alaska folds) versus **fold-mean** (each
of the `r prov$n_folds` locales weighted equally, so the tiny cities count as much
as Massachusetts).

```{r pooled-vs-foldmean}
metrics %>%
  select(model, oos_dev_r2, mae) %>%
  left_join(
    fold_mean %>%
      mutate(model = metrics$model[match(model, c("M1","M2","M3"))]),
    by = "model"
  ) %>%
  transmute(
    Model = model,
    `Dev-R² (precinct-pooled)` = round(oos_dev_r2, 3),
    `Dev-R² (fold-mean)`       = round(dev_r2_fold_mean, 3),
    `MAE pp (precinct-pooled)` = round(100 * mae, 2),
    `MAE pp (fold-mean)`       = round(100 * mae_fold_mean, 2)
  ) %>%
  kable(caption = "Precinct-pooled vs equal-weight-per-locale.") %>%
  kable_styling(full_width = FALSE)
```

## Per-held-out-locale detail

Which jurisdictions are hard to predict from the others? This is the table that
makes the LOLO test concrete.

```{r fold-table}
fold_metrics %>%
  filter(model == "M1") %>%
  transmute(
    `Held-out locale` = locale,
    `N precincts` = n_precincts,
    `OOS dev-R²` = round(oos_dev_r2, 3),
    `MAE (pp)`   = round(100 * mae, 2),
    `Mean observed support` = round(mean_obs, 3),
    `Mean predicted`        = round(mean_pred, 3)
  ) %>%
  kable(caption = "M1 baseline, one row per held-out locale.") %>%
  kable_styling(full_width = FALSE)
```

```{r fold-plot, fig.width=8, fig.height=5}
plot_df <- fold_metrics %>%
  mutate(model = recode(model, M1 = "M1 baseline",
                        M2 = "M2 + demographics", M3 = "M3 GLMM"))

ggplot(plot_df, aes(x = reorder(locale, mae), y = 100 * mae, fill = model)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  scale_fill_brewer(palette = "Dark2", name = NULL) +
  coord_flip() +
  labs(
    title = "Out-of-sample error by held-out locale",
    subtitle = "Mean absolute error of predicted RCV support share, leave-one-locale-out",
    x = NULL, y = "MAE (percentage points)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

```{r calib-plot, fig.width=7, fig.height=5.5}
# Observed vs predicted support, held-out precincts (M1), coloured by locale.
cal <- cv_results$pooled_preds$M1
ggplot(cal, aes(x = p, y = obs, color = locale)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(alpha = 0.35, size = 0.9) +
  scale_color_brewer(palette = "Set2", name = "Held-out locale") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title = "Held-out calibration (M1 baseline)",
    subtitle = "Each point is a precinct predicted while its locale was held out",
    x = "Predicted RCV support", y = "Observed RCV support"
  ) +
  theme_minimal(base_size = 12)
```

## Imputation uncertainty

For M2 / M3, how much does the choice of imputation move the headline numbers?
Small spread means the out-of-sample conclusion is not an artifact of any single
imputation.

```{r imp-spread}
imp_spread %>%
  transmute(
    Model = model,
    `Dev-R² mean` = round(dev_r2_mean, 3),
    `Dev-R² SD`   = round(dev_r2_sd, 3),
    `MAE mean (pp)` = round(100 * mae_mean, 2),
    `MAE SD (pp)`   = round(100 * mae_sd, 2),
    `Imputations` = m
  ) %>%
  kable(caption = paste0("Spread of headline metrics across the ",
                         prov$m_imputations,
                         " individual imputations (M1 has none).")) %>%
  kable_styling(full_width = FALSE)
```

## How to cite these numbers

```{r provenance}
tibble(
  Field = c("CV design", "Folds", "Precincts", "Imputations", "Null model",
            "Pooling", "Generated", "R", "mice", "git"),
  Value = c(prov$cv_design, prov$n_folds, prov$n_precincts, prov$m_imputations,
            prov$null_model, prov$pooling, prov$generated_at, prov$r_version,
            prov$mice_version, prov$git_sha)
) %>%
  kable(caption = "Provenance of the numbers in this report.") %>%
  kable_styling(full_width = FALSE)
```

Numbers are persisted to `data/cv_results.RData`, `data/cv_metrics.csv`, and
`data/cv_fold_metrics.csv`, regenerated by `R/05_cv.R`.
