---
title: "Low Plurality Winner (LPW) Data Sources"
author: "Jesse Brandt"
execute:
  echo: true
  warning: false
  message: false
---

## Background

The `recent_lpw` variable flags whether a locale's state had a statewide race
(governor or U.S. senator) won with less than 40% of the vote within 10 years
prior to the RCV ballot measure. The two cases in this study:

- **Maine**: Governor Paul LePage won in 2010 with 37.6%
  ([Ballotpedia](https://ballotpedia.org/Maine_gubernatorial_election,_2010))
- **Alaska**: Senator Lisa Murkowski won in 2010 with 39.7% (write-in)
  ([Ballotpedia](https://ballotpedia.org/Lisa_Murkowski))

The winner-share table behind `recent_lpw` is **derived programmatically**
from downloadable returns (fetched by `R/00_download.R`, derived in the
`recent_lpw` section of `R/01_clean_and_merge.R`):

| Races | Source |
|---|---|
| U.S. Senate, all 6 states | MEDSL U.S. Senate 1976--2020 (statewide totals) |
| Governor, 5 county states | Amlani & Algara county returns 1865--2020, summed to state |
| Governor, Alaska 2014 & 2018 | Official AK statewide summaries (Internet Archive) |
| Governor, Alaska 2010 | Hand-entered (`statewide_elections_manual.csv`) — no machine-readable source exists |

An earlier version of this project used a fully hand-extracted table
(`data/raw/statewide_elections_lpw_2006_2018.csv`, copied from Ballotpedia and
Wikipedia). That file is retained only as a cross-check: the pipeline stops if
the derived LPW classification ever disagrees with it.

## U.S. Senate returns (MEDSL)

- **Citation**: MIT Election Data and Science Lab, 2017, "U.S. Senate 1976--2020",
  <https://doi.org/10.7910/DVN/PEJ5QU>, Harvard Dataverse
- **File**: `data/raw/medsl_senate_1976_2020.csv`
- **Download**: Harvard Dataverse API, file ID 7609736 (see `R/00_download.R`)

### Senate races won with < 40% (2006+)

```{r senate-lpw}
library(dplyr)

sen <- read.csv("../data/raw/medsl_senate_1976_2020.csv")

senate_winners <- sen %>%
  filter(stage == "gen", writein == "False") %>%
  group_by(year, state, state_po, totalvotes) %>%
  summarise(winner_votes = max(candidatevotes), .groups = "drop") %>%
  mutate(winner_share = round(100 * winner_votes / totalvotes, 1))

senate_winners %>%
  filter(winner_share < 40, year >= 2006) %>%
  arrange(winner_share) %>%
  select(year, state, winner_share) %>%
  knitr::kable(caption = "U.S. Senate races won with < 40% (2006+)")
```

Alaska 2010 appears at 35.6% — this is Joe Miller's plurality among
non-write-in candidates. Murkowski's write-in share was 39.7%. Either way,
no candidate cleared 40%, confirming the LPW classification.

## Governor returns (Amlani & Algara)

MEDSL does not publish a consolidated governor returns file, but a
county-level gubernatorial returns dataset covering 1865--2020 is available
on Harvard Dataverse and aggregates cleanly to statewide totals.

- **Citation**: Amlani, Sharif; Algara, Carlos, 2021, Replication Data for
  "Partisanship & Nationalization in American Elections: Evidence from
  Presidential, Senatorial, & Gubernatorial Elections in the U.S. Counties,
  1872--2020" (*Electoral Studies*),
  <https://doi.org/10.7910/DVN/DGUMFI>, Harvard Dataverse
- **File**: `data/raw/gubernatorial_county_returns_1865_2020.RData`
  (loads as `gov_elections_release`)
- **Download**: Harvard Dataverse API, file ID 5028535 (see `R/00_download.R`)

Two limitations, both handled in the pipeline:

1. **Alaska is absent** — the state has no counties. Alaska governor races
   come from official Division of Elections files instead (next section).
2. Only Democratic and Republican raw votes are itemized (plus an
   all-candidate total), so winner share is computed as `max(D, R) / total`.
   This is correct whenever the winner was a D or R — true for every governor
   race in the five county states during the study window. Independent
   *losers* (e.g. Cutler, Maine 2010) are in the denominator and only shrink
   the computed winner share, the conservative direction for a < 40% flag.

```{r governor-shares}
load("../data/raw/gubernatorial_county_returns_1865_2020.RData")

state_po_map <- c(CA = "California", CO = "Colorado", ME = "Maine",
                  MA = "Massachusetts", MN = "Minnesota")

gov_winners <- gov_elections_release %>%
  filter(state %in% names(state_po_map), election_year >= 2006) %>%
  group_by(state, election_year) %>%
  summarise(dem = sum(democratic_raw_votes, na.rm = TRUE),
            rep = sum(republican_raw_votes, na.rm = TRUE),
            total = sum(raw_county_vote_totals, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(winner_share = round(100 * pmax(dem, rep) / total, 1))

gov_winners %>%
  select(state, election_year, winner_share) %>%
  arrange(state, election_year) %>%
  knitr::kable(caption = "Governor winner shares, study states (2006+)")
```

Maine 2006 (38.1%) and 2010 (38.1%) are the only sub-40% governor races —
matching the hand-extracted table (38.11% and 37.6%; the small 2010
difference reflects a different vote-total base, on the same side of 40%).

## Alaska governor races

The Alaska Division of Elections publishes machine-readable statewide results
summaries, but the live site now sits behind a CAPTCHA, so `R/00_download.R`
fetches the Internet Archive's captures of the official `results.txt` files:

- **2014**: `data/raw/alaska_statewide_results_2014.txt`
  (Walker/Mallott, an independent ticket, won with 48.1% — a case the
  two-party county dataset could not have handled)
- **2018**: `data/raw/alaska_statewide_results_2018.txt`
  (Dunleavy won with 51.4%)

**2010 is the one race with no machine-readable source anywhere.** The
official 2010 results file — in every archived form (txt, htm, pdf) — contains
only the U.S. Senate race (the Miller/Murkowski write-in count), and
OpenElections' Alaska coverage starts in 2014. The official record for the
rest of the 2010 ballot is the set of per-district SOVC PDFs. This single row
(Parnell, R, 59.06% — far above the 40% threshold) is hand-entered in
`data/raw/statewide_elections_manual.csv` with its source documented inline.

## Cross-check against the hand-extracted table

```{r cross-check}
hand <- read.csv("../data/raw/statewide_elections_lpw_2006_2018.csv",
                 fileEncoding = "UTF-8-BOM")

hand %>%
  filter(Winner.Share.... < 40) %>%
  select(State, Year, Office, Winner.Share....) %>%
  knitr::kable(caption = "Hand-extracted table: races won with < 40%")
```

Both the derived table and the hand-extracted table classify exactly
**Maine and Alaska** as LPW states. `R/01_clean_and_merge.R` asserts this
agreement with `stopifnot()` on every run.

## Expanding to new locales

Adding a state now requires no hand extraction for any senate race
(MEDSL, through 2020) or any governor race in a county state (Amlani &
Algara, through 2020). Newer races would need the successor MEDSL releases
or state sources; genuinely unsourceable races go in
`statewide_elections_manual.csv` with a documented source.
