---
title: "Low Plurality Winner (LPW) Data Sources"
author: "Jesse Brandt"
execute:
  echo: true
  warning: false
  message: false
---

## Background

The `recent_lpw` variable flags whether a locale's state had a statewide race
(governor or U.S. senator) won with less than 40% of the vote within 10 years
of the RCV ballot measure. The two cases in this study:

- **Maine**: Governor Paul LePage won in 2010 with 37.6%
  ([Ballotpedia](https://ballotpedia.org/Maine_gubernatorial_election,_2010))
- **Alaska**: Senator Lisa Murkowski won in 2010 with 39.7% (write-in)
  ([Ballotpedia](https://ballotpedia.org/Lisa_Murkowski))

The current project uses a hand-built CSV
(`data/raw/statewide_elections_lpw_2006_2018.csv`) covering the 6 states in
the dataset. This is sufficient for the current scope.

## U.S. Senate returns (MEDSL)

For reference, a consolidated U.S. Senate general election returns file
(1976--2020) from the MIT Election Data + Science Lab has been downloaded.

- **Citation**: MIT Election Data and Science Lab, 2017, "U.S. Senate 1976--2020",
  <https://doi.org/10.7910/DVN/PEJ5QU>, Harvard Dataverse
- **File**: `data/raw/medsl_senate_1976_2020.csv`
- **Download**: Harvard Dataverse API, file ID 7609736

```{r download-senate, eval=FALSE}
download.file(
  url = "https://dataverse.harvard.edu/api/access/datafile/7609736?format=original",
  destfile = "data/raw/medsl_senate_1976_2020.csv",
  mode = "wb"
)
```

### Senate races won with < 40% (2006+)

```{r senate-lpw}
library(dplyr)

sen <- read.csv("../data/raw/medsl_senate_1976_2020.csv")

senate_winners <- sen %>%
  filter(stage == "gen", writein == "False") %>%
  group_by(year, state, state_po, totalvotes) %>%
  summarise(winner_votes = max(candidatevotes), .groups = "drop") %>%
  mutate(winner_share = round(100 * winner_votes / totalvotes, 1))

senate_winners %>%
  filter(winner_share < 40, year >= 2006) %>%
  arrange(winner_share) %>%
  select(year, state, winner_share) %>%
  knitr::kable(caption = "U.S. Senate races won with < 40% (2006+)")
```

Alaska 2010 appears at 35.6% — this is Joe Miller's plurality among
non-write-in candidates. Murkowski's write-in share was 39.7%. Either way,
no candidate cleared 40%, confirming the LPW classification.

## Future research

MEDSL does not publish a consolidated governor returns file. If expanding to
new locales, governor races would need to be checked manually (e.g. from
Ballotpedia) or scraped. Low-plurality governors are rare; for the current
8-locale dataset the hand-built CSV is sufficient.
