---
title: "RCV Ballot Measure Models"
author: "Jesse Brandt"
format: html
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(lme4)
library(scales)
library(broom)
library(broom.mixed)
library(knitr)

load("../data/rcv_data.RData")
```

## Data Preparation

The cleaning pipeline (`R/01_clean_and_merge.R`) now produces unified
`rcv_yes` and `rcv_no` columns across all 8 locales.

```{r prep}
df <- states_and_cities %>%
  filter(!is.na(rcv_yes), !is.na(rcv_no), !is.na(dem_share),
         (rcv_yes + rcv_no) > 0) %>%
  mutate(
    n_votes   = rcv_yes + rcv_no,
    recent_lpw = factor(recent_lpw, levels = c(FALSE, TRUE),
                        labels = c("No", "Yes"))
  )

df %>%
  group_by(locale, rcv_jurisdiction, recent_lpw) %>%
  summarise(n_precincts = n(), total_votes = sum(n_votes),
            mean_yes = mean(yes_share), .groups = "drop") %>%
  kable(digits = 3)
```

## 1 -- Binomial GLM (pooled)

A binomial GLM with a logit link uses the actual vote counts as the response,
properly accounting for the bounded [0, 1] nature of `yes_share` and
giving more weight to precincts with more voters.

```{r glm-binomial}
m_binom <- glm(
  cbind(rcv_yes, rcv_no) ~ dem_share + recent_lpw,
  family = binomial(link = "logit"),
  data = df
)

summary(m_binom)
```

The deviance-to-df ratio measures overdispersion (>>1 means the binomial
variance assumption is too tight -- common with aggregate vote data).

```{r overdispersion}
cat("Residual deviance / df:",
    round(deviance(m_binom) / df.residual(m_binom), 1), "\n")
```

Overdispersion is expected: precinct vote shares vary more than pure binomial
sampling because voters within a precinct are not independent coin flips.
A **quasibinomial** model uses the same point estimates but corrects the
standard errors.

```{r glm-quasi}
m_quasi <- glm(
  cbind(rcv_yes, rcv_no) ~ dem_share + recent_lpw,
  family = quasibinomial(link = "logit"),
  data = df
)

tidy(m_quasi, conf.int = TRUE) %>%
  kable(digits = 4, caption = "Quasibinomial GLM: yes ~ dem_share + recent_lpw")
```

### Adding the interaction (dem_share x recent_lpw)

Does the effect of partisanship on RCV support differ in locales with a
recent low-plurality winner?

```{r glm-interaction}
m_quasi_int <- glm(
  cbind(rcv_yes, rcv_no) ~ dem_share * recent_lpw,
  family = quasibinomial(link = "logit"),
  data = df
)

tidy(m_quasi_int, conf.int = TRUE) %>%
  kable(digits = 4,
        caption = "Quasibinomial GLM with interaction")
```

```{r glm-compare}
anova(m_quasi, m_quasi_int, test = "F")
```

### Marginal effects plot

```{r glm-fitted, fig.width=8, fig.height=5}
pred_grid <- expand.grid(
  dem_share  = seq(0, 1, length.out = 200),
  recent_lpw = factor(c("No", "Yes"))
)

pred_grid$fit <- predict(m_quasi_int, newdata = pred_grid, type = "response")

ggplot() +
  geom_point(data = df,
             aes(x = dem_share, y = yes_share, color = recent_lpw),
             alpha = 0.3, size = 0.8) +
  geom_line(data = pred_grid,
            aes(x = dem_share, y = fit, color = recent_lpw),
            linewidth = 1.2) +
  scale_x_continuous(labels = percent_format(1)) +
  scale_y_continuous(labels = percent_format(1), limits = c(0, 1)) +
  scale_color_manual(values = c("No" = "#4575B4", "Yes" = "#D73027")) +
  labs(x = "Democratic Presidential Vote Share",
       y = "RCV Yes Vote Share",
       color = "Recent low-\nplurality winner",
       title = "Quasibinomial GLM with interaction") +
  theme_bw(base_size = 13)
```

## 2 -- Mixed-effects binomial model (random intercepts by locale)

Precincts are nested within locales that differ in unobserved ways
(campaign spending, ballot language, media environment). A random-intercept
model accounts for this clustering.

```{r glmer-ri}
m_glmer <- glmer(
  cbind(rcv_yes, rcv_no) ~ dem_share + recent_lpw + (1 | locale),
  family = binomial,
  data = df,
  nAGQ = 10
)

summary(m_glmer)
```

```{r glmer-ri-table}
tidy(m_glmer, conf.int = TRUE, effects = "fixed") %>%
  kable(digits = 4, caption = "GLMM: random intercept by locale")
```

Random-intercept estimates (locale-level deviations on the log-odds scale):

```{r ranef}
ranef(m_glmer)$locale %>%
  tibble::rownames_to_column("locale") %>%
  rename(intercept_shift = `(Intercept)`) %>%
  arrange(desc(intercept_shift)) %>%
  kable(digits = 3)
```

### With interaction

```{r glmer-int}
m_glmer_int <- glmer(
  cbind(rcv_yes, rcv_no) ~ dem_share * recent_lpw + (1 | locale),
  family = binomial,
  data = df,
  nAGQ = 10
)

tidy(m_glmer_int, conf.int = TRUE, effects = "fixed") %>%
  kable(digits = 4,
        caption = "GLMM with interaction + random intercept")
```

```{r glmer-compare}
anova(m_glmer, m_glmer_int)
```

### Random slopes

Does the effect of `dem_share` vary by locale? Adding a random slope
lets each locale have its own partisan gradient.

```{r glmer-rs}
m_glmer_rs <- glmer(
  cbind(rcv_yes, rcv_no) ~ dem_share * recent_lpw + (1 + dem_share | locale),
  family = binomial,
  data = df,
  nAGQ = 1   # Laplace; nAGQ > 1 not supported with random slopes
)

summary(m_glmer_rs)
```

```{r glmer-rs-ranef}
coef(m_glmer_rs)$locale %>%
  tibble::rownames_to_column("locale") %>%
  kable(digits = 4, caption = "Locale-specific coefficients (random slopes)")
```

### Fitted curves from the random-slopes model

```{r glmer-rs-plot, fig.width=9, fig.height=6}
locale_coefs <- coef(m_glmer_rs)$locale %>%
  tibble::rownames_to_column("locale")

locale_meta <- df %>%
  distinct(locale, recent_lpw)

pred_rs <- locale_coefs %>%
  left_join(locale_meta, by = "locale") %>%
  tidyr::crossing(dem_share_grid = seq(0, 1, length.out = 200)) %>%
  mutate(
    lpw_val = ifelse(recent_lpw == "Yes", 1, 0),
    eta = `(Intercept)` + dem_share * dem_share_grid +
          `recent_lpwYes` * lpw_val +
          `dem_share:recent_lpwYes` * dem_share_grid * lpw_val,
    fitted = plogis(eta)
  )

ggplot() +
  geom_point(data = df,
             aes(x = dem_share, y = yes_share, color = locale),
             alpha = 0.25, size = 0.7) +
  geom_line(data = pred_rs,
            aes(x = dem_share_grid, y = fitted, color = locale),
            linewidth = 1) +
  scale_x_continuous(labels = percent_format(1)) +
  scale_y_continuous(labels = percent_format(1), limits = c(0, 1)) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Democratic Presidential Vote Share",
       y = "RCV Yes Vote Share",
       color = "Locale",
       title = "Mixed-effects binomial model (random slopes by locale)") +
  theme_bw(base_size = 13)
```

## 3 -- Model comparison summary

AIC is undefined for quasibinomial models, so the comparison below includes
only the GLMM fits (which have proper likelihoods). The quasibinomial models
are compared via the F-test above.

```{r comparison}
models <- tibble(
  Model = c(
    "GLMM random intercept",
    "GLMM random intercept + interaction",
    "GLMM random slopes + interaction"
  ),
  AIC = c(AIC(m_glmer), AIC(m_glmer_int), AIC(m_glmer_rs))
)

models %>%
  mutate(AIC = format(round(AIC), big.mark = ",")) %>%
  kable(caption = "AIC comparison (lower is better)")
```
