---
title: "RCV Models with Demographic Controls (Imputed Data)"
author: "Jesse Brandt"
format: html
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
library(mice)
library(dplyr)
library(lme4)
library(broom)
library(broom.mixed)
library(knitr)
library(kableExtra)
library(ggplot2)
library(scales)

load("../data/model_results.RData")
```

## Overview

All models use a **quasibinomial GLM** with logit link on `cbind(rcv_yes, rcv_no)` —
each ballot is treated as a Bernoulli trial, with quasibinomial correction for
overdispersion. Results are pooled across 5 multiply-imputed datasets using Rubin's rules.
Imputation excluded all RCV outcome variables.

**Models:**

- **M1**: Quasibinomial baseline — `dem_share + recent_lpw` only
- **M2**: Quasibinomial + demographic controls
- **M3**: GLMM with random intercept by locale + controls

**Demographic controls** (PL94-171 + ACS 5-year):
`pct_white_vap`, `pct_black_vap`, `pct_hispanic_vap`, `median_income`, `pct_bach_plus`, `pct_renter`

## Data summary

```{r summary}
diag_data %>%
  group_by(locale, rcv_jurisdiction, recent_lpw) %>%
  summarise(
    n = n(),
    mean_yes = mean(yes_share),
    mean_dem = mean(dem_share),
    mean_white = mean(pct_white_vap),
    med_income = median(median_income),
    mean_bach = mean(pct_bach_plus),
    .groups = "drop"
  ) %>%
  kable(digits = 3, caption = "Analysis sample by locale")
```

## Model 1: Quasibinomial baseline

$$\text{logit}(p) = \beta_0 + \beta_1 \cdot \text{dem\_share} + \beta_2 \cdot \text{recent\_lpw}$$

```{r m1}
summary(pool_m1) %>%
  kable(digits = 4, caption = "M1: Quasibinomial baseline (pooled)")
```

## Model 2: Quasibinomial with demographic controls

$$\text{logit}(p) = \beta_0 + \beta_1 \cdot \text{dem\_share} + \beta_2 \cdot \text{recent\_lpw} + \boldsymbol{\gamma} \cdot \mathbf{X}_{\text{demo}}$$

```{r m2}
summary(pool_m2) %>%
  kable(digits = 4, caption = "M2: Quasibinomial + controls (pooled)")
```

### Adjusted pseudo-R² comparison

Deviance-based: $1 - \frac{\text{deviance}/\text{df}_{\text{resid}}}{\text{null deviance}/\text{df}_{\text{null}}}$. Penalizes for number of parameters.

```{r pseudo-r2}
tibble(
  Model = c("M1: Baseline", "M2: + Demographics"),
  `Pseudo-R²` = round(c(pr2_m1$pseudo_r2, pr2_m2$pseudo_r2), 4),
  `Adj. Pseudo-R²` = round(c(pr2_m1$adj_pseudo_r2, pr2_m2$adj_pseudo_r2), 4)
) %>%
  kable(caption = "Deviance-based pseudo-R² (averaged across 5 imputations)")
```

### F-test: are demographic controls jointly significant?

```{r f-test}
f_tests <- lapply(1:length(fit_m1$analyses), function(i) {
  anova(fit_m1$analyses[[i]], fit_m2$analyses[[i]], test = "F")
})

tibble(
  Imputation = 1:length(f_tests),
  `F statistic` = sapply(f_tests, function(a) round(a$F[2], 1)),
  `p-value` = sapply(f_tests, function(a) format.pval(a$`Pr(>F)`[2], digits = 3))
) %>%
  kable(caption = "F-test: M1 vs M2 (demographic controls jointly)")
```

### Overdispersion

```{r overdispersion}
tibble(
  Model = c("M1: Baseline", "M2: + Controls"),
  `Dispersion (dev/df)` = c(
    round(mean(sapply(fit_m1$analyses, function(m) deviance(m)/df.residual(m))), 1),
    round(mean(sapply(fit_m2$analyses, function(m) deviance(m)/df.residual(m))), 1)
  )
) %>%
  kable(caption = "Overdispersion (1.0 = none; quasibinomial corrects SEs)")
```

### Interpretation (M2)

All coefficients are on the **log-odds scale**. The core finding holds: `dem_share`
and `recent_lpw` remain strong positive predictors of RCV support. Among the
demographic controls:

- **`pct_bach_plus`** (education) is the strongest new predictor — more educated
  precincts support RCV more, controlling for partisanship.
- **`pct_renter`** is positive — more urban/transient precincts favor RCV.
- **`pct_black_vap`** is strongly negative — precincts with more Black residents
  are less supportive, holding partisanship constant.
- **`median_income`** is negative — wealthier precincts slightly less supportive,
  holding education constant.

## Model 3: GLMM with random intercept by locale

Accounts for unobserved locale-level variation (campaign spending, ballot language,
media environment) via random intercepts.

```{r m3}
summary(pool_m3) %>%
  kable(digits = 4, caption = "M3: GLMM random intercept + controls (pooled)")
```

### Random effects (imputation 1)

```{r m3-ranef}
ranef(fit_m3$analyses[[1]])$locale %>%
  tibble::rownames_to_column("locale") %>%
  rename(intercept_shift = `(Intercept)`) %>%
  arrange(desc(intercept_shift)) %>%
  kable(digits = 3,
        caption = "Locale random intercepts (log-odds, imputation 1)")
```

## Model comparison

```{r comparison}
tibble(
  Model = c("M1: Quasibinom baseline",
            "M2: Quasibinom + demographics",
            "M3: GLMM + demographics"),
  `Adj. pseudo-R²` = c(
    round(pr2_m1$adj_pseudo_r2, 4),
    round(pr2_m2$adj_pseudo_r2, 4),
    "—"
  ),
  Notes = c(
    "",
    "Demographics jointly significant (F-test p < 2e-16)",
    "Random intercept absorbs locale-level heterogeneity"
  )
) %>%
  kable(caption = "Model overview")
```

## Coefficient comparison: M1 vs M2

```{r coefplot-m1m2, fig.width=9, fig.height=6}
coefs <- bind_rows(
  summary(pool_m1) %>% mutate(model = "M1: Baseline"),
  summary(pool_m2) %>% mutate(model = "M2: + Demographics")
) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    lo = estimate - 1.96 * std.error,
    hi = estimate + 1.96 * std.error,
    term = case_match(term,
      "dem_share" ~ "Dem. vote share",
      "recent_lpw" ~ "Recent LPW (=TRUE)",
      "pct_white_vap" ~ "% White VAP",
      "pct_black_vap" ~ "% Black VAP",
      "pct_hispanic_vap" ~ "% Hispanic VAP",
      "median_income" ~ "Median income",
      "pct_bach_plus" ~ "% Bachelor's+",
      "pct_renter" ~ "% Renter",
      .default = term
    )
  )

ggplot(coefs, aes(x = estimate, y = reorder(term, estimate), color = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = lo, xmax = hi),
                  position = position_dodge(width = 0.5), size = 0.5) +
  scale_color_manual(values = c("M1: Baseline" = "#4575B4",
                                "M2: + Demographics" = "#D73027")) +
  labs(x = "Estimate (log-odds, 95% CI)", y = NULL, color = NULL,
       title = "Quasibinomial coefficient comparison") +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

## Coefficient comparison: M2 vs M3

```{r coefplot-m2m3, fig.width=10, fig.height=6}
coefs_all <- bind_rows(
  summary(pool_m2) %>% mutate(model = "M2: Quasibinom"),
  summary(pool_m3) %>% mutate(model = "M3: GLMM")
) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    lo = estimate - 1.96 * std.error,
    hi = estimate + 1.96 * std.error,
    term = case_match(term,
      "dem_share" ~ "Dem. vote share",
      "recent_lpw" ~ "Recent LPW (=TRUE)",
      "pct_white_vap" ~ "% White VAP",
      "pct_black_vap" ~ "% Black VAP",
      "pct_hispanic_vap" ~ "% Hispanic VAP",
      "median_income" ~ "Median income",
      "pct_bach_plus" ~ "% Bachelor's+",
      "pct_renter" ~ "% Renter",
      .default = term
    )
  )

ggplot(coefs_all, aes(x = estimate, y = reorder(term, estimate), color = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = lo, xmax = hi),
                  position = position_dodge(width = 0.5), size = 0.5) +
  scale_color_manual(values = c("M2: Quasibinom" = "#4575B4",
                                "M3: GLMM" = "#1B7837")) +
  labs(x = "Estimate (log-odds, 95% CI)", y = NULL, color = NULL,
       title = "Quasibinomial vs. GLMM coefficient comparison") +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```
